---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data
```{r}
library(dplyr)
library(ggplot2)
library(tidyr)
activity <- read.csv("activity.csv", header = TRUE)
activity$date <- as.Date(as.character(activity$date), format = "%Y-%m-%d")
activity$steps <- as.numeric(as.character(activity$steps))
activity$interval <- as.numeric(as.character(activity$interval))

```
## What is mean total number of steps taken per day?
# NA values were removed, so any days with NA values were not included in the following figure
*The 'zeroes' column is probably accounted for by the NA values.*
```{r}

steps_per_day <- activity %>% group_by(date) %>% summarise(sum(steps, na.rm = TRUE))

hist(steps_per_day$`sum(steps, na.rm = TRUE)`, breaks = 10, xlim = c(0, 25000),
    xlab = "Total Steps", main = "Histogram of Total Steps Taken each Day
    October-November 2012")

mean_spd <- mean(steps_per_day$`sum(steps, na.rm = TRUE)`)
median_spd <- median(steps_per_day$`sum(steps, na.rm = TRUE)`)

```

The mean number of steps taken each day, ignoring missing values was `r mean_spd` and the median was `r median_spd`.

## What is the average daily activity pattern?
```{r}

avg_interval_steps <- activity %>% group_by(interval) %>% summarise(mean(steps, na.rm = TRUE))
names(avg_interval_steps) <- c("interval", "average steps")

g <- ggplot(avg_interval_steps, aes(interval, `average steps`) )
g  + geom_line(col = "blue", lwd = .3) + xlab("5 Minute Intervals (Over 24 hrs.) ") + ylab("Average Number of Steps")+
    labs(title = "Steps Taken for each 5 Minute Interval in 24 Hours ", subtitle = "Average for Equivalent 5 Minute Intervals over 2 Months")

max_steps <- avg_interval_steps[avg_interval_steps$`average steps`==max(avg_interval_steps$`average steps`),]
max_steps_interval <- max_steps[1,1] 
max_steps <- max_steps[1,2]

```

On average, across all of the days in the dataset, the 5-minute interval at `r max_steps_interval` contains the maximum number of steps, `r max_steps`.

## Imputing missing values

```{r}
mv_count <- sum(is.na(activity$steps))

#replace NAs with mean average for that 5 minute interval
activity_nmv <- activity # new dataset with replaced NAs
steps_vector <- activity_nmv$steps

for (i in 1:length(steps_vector)){
    
   if(is.na(steps_vector[i])){
       
      interval_val <- activity_nmv$interval[i]
      
    mean_val <- subset(avg_interval_steps, avg_interval_steps$interval == interval_val)
    mean_val <- as.numeric(mean_val[1,2])
    
    activity_nmv$steps[i] <- coalesce(activity_nmv$steps[i], mean_val) 
   }
    
    
} 

steps_no_NA <- activity_nmv %>% group_by(date) %>% summarise(sum(steps))

hist(steps_no_NA$`sum(steps)`, breaks = 10, xlim = c(0, 25000),
    xlab = "Total Steps", main = "Histogram of Total Steps Taken each Day
    October-November 2012 (Imputed NA values)", col = "red")

mean_new <- mean(steps_no_NA$`sum(steps)`)
median_new <- median(steps_no_NA$`sum(steps)`)

```

There are `r mv_count` missing values in the dataset.(i.e. total number of rows with NAs)

The mean total steps each day is now `r mean_new` and the updated median is `r median_new`.

Imputing the NA values has the effect of increasing both the mean average and the median as any rows
with NA values were previously evaluated to zero, greatly increasing the proportion of days where zero 
steps were taken which wasn't accurate.

## Are there differences in activity patterns between weekdays and weekends?

```{r}
weekdays1 <- c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday')
activity_nmv$wkday_wkend <- factor((weekdays(activity_nmv$date) %in% weekdays1), 
         levels=c(FALSE, TRUE), labels=c('weekend', 'weekday'))

wkday_wkend_steps <- activity_nmv %>% group_by(wkday_wkend,interval) %>% summarise(mean(steps))
names(wkday_wkend_steps) <- c("weekday/weekend", "interval", "average steps")

g <- ggplot(wkday_wkend_steps, aes(interval, `average steps`) )
g  + geom_line(col = "blue", lwd = .3) + facet_wrap(wkday_wkend_steps$`weekday/weekend`, nrow=2, ncol = 1)+
    xlab("5 Minute Intervals (Over 24 hrs.) ") + ylab("Average Number of Steps")+
    labs(title = "Steps Taken for each 5 Minute Interval in 24 Hours ", subtitle = "Average for Equivalent 5 Minute Intervals over 2 Months")

```
